### 2020-11-14 docs/manuscript.tex
add 'Issues with using p-values as a screening tool' subsection

\PaperTitle{
    HAMDetector: Combining information to predict HLA-associated mutations with a
    Bayesian regression model
} % Article title

@@

\section{System and methods}
\section{Algorithm}
\section{Implementation}
\section{Discussion}
\section{References}

>>>>

\PaperTitle{
    HAMDetector: Combining information to detect HLA-associated mutations with a
    Bayesian regression model
} % Article title

@@

\subsection{Computational methods}

Because the currently available experimental methods do not scale well enough to analyze
whole viral genomes, a useful strategy might be to use annotated sequence data to identfiy 
candidate HLA escape mutations that can then be verified experimentally.

As the selection pressure exerted by cytotoxic T cells depends on successful recognition
of viral peptides on the cell surface, escape mutations are often HLA-allele specific
and can therefore  be detected as HLA-allele dependent footprints in sequence
alignments of viral proteins: At certain alignment positions, a replacement
might be more frequently observed in sequences from hosts with a specific HLA allele than in 
sequences from hosts without that HLA allele. By quantifying this difference
for all replacement and HLA allele pairs it is possible to idenfify replacements that
are enriched in sequences coming from hosts with certain HLA alleles and are therefore
likely to be HLA escape mutations.

One way of quantifying this enrichment is Fisher\'s exact test.
For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\),
a 2-by-2 contingency table is constructed containing the absolute counts of the
number of sequences in the four possible categories: 
(\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)),
where \(!R_{i}\) denotes any replacement except \(R_{i}\) and \(!H\) denotes any HLA allele
except \(H\).
Fisher\'s exact test is a conventional nullhypothesis significance test (NHST) that
generates p-values. In this case, the nullhypothesis is that HLA allele \(H\) and 
replacement \(R_{i}\) are independent, and the p-value is the probability of
observing a deviation from independence that is at least as extreme as in the data at hand,
under the assumption that the nullhypothesis is true.

Fisher\'s exact test has the advantage of being easy to apply, but has also several 
disadvantages that are outlined in Carlson et al.
The most striking one is that viral sequences share a common phylogenetic history and therfore,
treating sequences as independent and identically distributed samples may under- or
overestimate effect sizes and in the context of hypothesis testing leads to
increased false positive and false negative rates.
Another issue of applying Fisher\'s exact test is that HLA class I loci are located in
close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
that HLA alleles are not inherited completely independent of each other, so that
inheritance of one HLA allele correlates with inheritance of another HLA allele.
When using a statistical method that tests each HLA allele individually without
considering the whole set of alleles present, spurious associations might occur
: If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but
\(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that
we observe an association between replacement \(R\) and \(H_{2}\), even in
absence of any underlying escape mechanism.

Correlations can not only occur between HLA alleles, but also between replacements.
This kind of codon covariation occurs for example in compensatory mutations that
attenuate the negative impact of immune escape mutations. For instance,
a compensatory mutation might lead to a conformational change in such a way that
the mutated protein resembles the original wild type more closely.

Carlson et. al. developed a method called Phylogenetic Dependency Network
that accounts for phylogenetic bias, HLA linkage disequilibrium and codon covariation.
and is based on nullhypothesis significance tests.

\subsection{Issues with using p-values as a screening tool}

In addition to the aforementioned biological reasons why Fisher's exact test is 
not suited for the analysis of sequence data, there are also more fundamental statistical
issues that universally occur when using p-values as a screening tool.

In the presence of small effect sizes and high variance between measurements, as 
it is typically the case when working with biological data, statistically significant
results can often be misleading and are likely to be in the wrong direction
(a so-called type S error) or greatly overestimate an effect (a so-called type M error).
This problem has recently been widely appreciated in the literature in the context
of the current \glqq replication crisis\grqq{}, which describes the circumstance that
scientific claims with seemingly strong statistical evidence fail to replicate.

Another issue that occurs when using p-values as a screening tool is the problem
of multiple comparisons. When applying a statistical test, the probability  of 
obtaining a statistically significant result increases with
each additional test, even in absence of any real effect. When using p-values as a filter
it is therefore likely to obtain significant effects that are in fact not real.
To circumvent this problem, a common strategy is to control the false discovery
rate, which is the expected proportion of false positives.
These adjustment procedures have the problem that, when performing many of such comparisons,
none but the very largest effects remain.
Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer
to fit a single, multilevel model that contains all comparisons of interest.
In this setting, the problem of multiple comparisons can disappear entirely and also
yield more valid estimates.

\section{System and methods}

\section{Algorithm}
\section{Implementation}
\section{Discussion}
\section{References}
